---
title: "Lab 9: Split plot designs"
subtitle: "FANR 6750: Experimental Methods in Forestry and Natural Resources Research"
author: "Fall 2021"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{lab09_splitplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.align = "c", fig.height = 4, fig.width = 4,
  fig.retina = 2,
  warning = FALSE, message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
source(here::here("R/zzz.R"))
library(kableExtra)
```

# Scenario

Remember from lecture that split-plot designs are used when we apply treatments to two experimental units: whole-units and sub-units

For example

- Ag fields are sprayed with herbicides, and fertilizers are applied to plots within fields  

- Tenderizer is applied to roasts, and cooking times are applied to cores  

In these cases, we’re interested in treatment effects at **both** levels

## The additive model

We can write the split plot model as:

$$\Large y_{ijk} = \mu + \alpha_i + \beta_j + \alpha \beta_{ij} + \gamma_k + \delta_{ik} + \epsilon_{ijk}$$
The $\alpha$’s and $\beta$’s are fixed treatment effects. Note the interaction.

Because we want our inferences to apply to all whole units (e.g., roasts), $\delta_{ik}$ is random. Specifically:

$$\Large \delta_{ik} \sim normal(0, \sigma^2_D)$$

We might view block (e.g. carcass) as random too:

$$\Large \gamma_k \sim normal(0, \sigma^2_C)$$

And as always,

$$\Large \epsilon_{ijk} \sim normal(0, \sigma^2)$$

## Example: The meat data

```{r setup}
library(FANR6750)
data("meatData")
str(meatData)
```

As we've done with other examples, we need to convert each predictor to a factor (in this case, converting to factors is necessary to use the `mcp()` function, later in the analysis):

```{r}
meatData$time <- factor(meatData$time)
meatData$carcass <- factor(meatData$carcass)
meatData$roast <- factor(meatData$roast)
meatData$tenderizer <- factor(meatData$tenderizer)
```

***

What does it mean to treat time as a factor? How would you interpret the results from a model that treated time as a continuous predictor vs a factor?

***

Notice that all the roast column does is enumerate the tenderizer × carcass combinations (which explains why the corresponding term in the additive model is $\delta_{ik}$):

```{r}
head(meatData, n=10)
```

## Carcass (block) effects as fixed

Only one `Error()` term allowed in `aov()`, so treat block effect as fixed:

```{r}
aov.meat <- aov(Wbscore ~ tenderizer * time + carcass + Error(roast), 
                data = meatData)
summary(aov.meat)
```

## Using `lme()` instead of `aov()`

```{r}
library(nlme)
lme.meat <- lme(Wbscore ~ tenderizer * time,
                data = meatData,
                correlation = corCompSymm(), # To make results same as aov()
                random = ~1|carcass/roast)
```

The first row of the ANOVA table is for the (Intercept) term, which we can ignore here (hence the `[-1,]` indexing operation):

```{r}
anova(lme.meat)[-1,]
```

### Exploring interaction

Is the time effect significant for each level of tenderizer?

```{r}
lme.meatP <- lme(Wbscore ~ time, data = meatData,
                 random = ~1|carcass/roast, 
                 correlation = corCompSymm(),
                 subset = tenderizer=="P")
anova(lme.meatP, Terms = "time")
```


```{r}
lme.meatV <- lme(Wbscore ~ time, data = meatData,
                 random = ~1|carcass/roast, 
                 correlation = corCompSymm(),
                 subset = tenderizer=="V")
anova(lme.meatV, Terms = "time")
```

```{r}
lme.meatC <- lme(Wbscore ~ time, data = meatData, 
                 random = ~1|carcass/roast, 
                 correlation = corCompSymm(),
                 subset = tenderizer=="C")
anova(lme.meatC, Terms = "time")
```

Yes it is.

## Multiple comparisons using `glht()` (package `multcomp`)

```{r}
library(multcomp)
mcP <- glht(lme.meatP, linfct = mcp(time="Tukey"))
summary(mcP)
```

We can also easily plot the differences between cooking times for “P” tenderizer (and for the other 2 levels of tenderizer as well)

```{r}
plot(mcP)
```

# Nested and crossed example

Nested-and-crossed is same as split-plot, but without the block. For example:

- Sweet potato yield is studied in response to ($a=3$) types of herbicide.

- Each herbicide is applied to 5 fields

- Each field is divided into 4 plots. Each plot is treated with one of ($b=4$) fertilizers.

## Assignment

The data can be accessed using:

```{r}
library(FANR6750)
data("yieldData")
```

1) Conduct the appropriate ANOVA using `aov()` and `lme()`

2) Give the appropriate additive model (this will be slightly different than the one for the split-plot design, since there is no blocking). What are the associated null and alternative
hypotheses?

3) Does the effect of herbicide depend on fertilizer?

4) Use Tukey’s test to determine which fertilizers differ

Put your answers in an R Markdown report. Your report should include:

1) A well-formatted ANOVA table (with caption); and

2) A publication-quality plot (or plots) of the estimated effect of herbicide and fertilizer on yield, including 95% confidence intervals. The figure should also have a descriptive caption and any aesthetics (color, line type, etc.) should be clearly defined.

You may format the report however you like but it should be well-organized, with relevant headers, plain text, and the elements described above. 

As always:

- Be sure the output type is set to: `output: html_document`

- Title the document: `title: "Lab 9 assignment"`

- Be sure to include your first and last name in the `author` section 

- Be sure to set `echo = TRUE` in all `R` chunks so we can see both your code and the output

- Please upload both the `html` and `.Rmd` files when you submit your assignment  

- See the R Markdown reference sheet for help with creating `R` chunks, equations, tables, etc.

