---
title: "Lab 2: t-tests"
subtitle: "FANR 6750: Experimental Methods in Forestry and Natural Resources Research"
author: "Fall 2021"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{lab02_t-tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.retina = 2,
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(knitr)
library(FANR6750)
library(WILD6900)
```

## Last week

- Using `R` as calculator

- Scripts

- Vectors and object classes

- Subsetting vectors

- Built-in functions/help

- Data frames

- Subsetting data frames


## Today's topics

1) Introduction

2) Graphics

3) *t*-tests

    + Two-sample t test  

    + Equality of variance test  

    + Paired t-test  
    
# Introduction

## Scenario

- We have 2 samples of data  

- **Question**: Do the samples come from the same population, or do they come from populations with different means?  

- **Problem**: We don’t know the true population means ($\mu_1$, $\mu_2$)  

- Under the assumption that the variances of the two populations are equal, the relevant hypotheses are:  

    + $H_0 : \mu_1 = \mu_2$  

    + $H_A : \mu_1 \neq \mu_2$

```{r echo = FALSE, fig.width=6, fig.density=3, fig.align='center'}
x <- seq(from = -5, to = 5, length.out = 1000)
y1 <- dnorm(x, -1, 1)
y2 <- dnorm(x, 1, 1)

df <- data.frame(x = rep(x, 2),
                 y = c(y1, y2),
                 pop = rep(c("A", "B"), each = 1000))

ggplot(df, aes(x = x, y = y, color = pop)) +
  geom_path() +
  guides(color = "none") +
  scale_x_continuous("") +
  scale_y_continuous("") +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  geom_segment(aes(x = -1, xend = -1, y = -Inf, yend = max(y1)), color = "black") +
  geom_segment(aes(x = 1, xend = 1, y = -Inf, yend = max(y1)), color = "black") +
  annotate("text", label = "bar(y)[1]", parse = TRUE, x = -1, y = max(y1) + 0.05) +
  annotate("text", label = "bar(y)[2]", parse = TRUE, x = 1, y = max(y1) + 0.05)
```

## Key points

- If the two sample means ($\bar{y}_1$, $\bar{y}_2$) are very different and the standard error of the difference in means is small, the $t$-statistic will be far from zero  

- If the $t$-statistic is more extreme than the critical values, we reject the null hypothesis ($H_0$)

```{r echo = FALSE, fig.width=6, fig.density=3, fig.align='center'}
op <- par(mai=c(0.8, 0.2, 0.2, 0.2))
curve(dt(x, df=18), -4, 4, xlab="t-value", ylab="", yaxt="n",
      ylim=c(0,0.4),
      frame=FALSE , main="t distribution with df=18", cex.main=1.5)
xs1 <- seq(qt(.025, df=18), -4, by=-0.1)
ys1 <- dt(xs1, df=19)
xs2 <- seq(qt(.975, df=18), 4, by=0.1)
ys2 <- dt(xs2, df=19)
polygon(c(xs1, rev(xs1)), c(rep(0, length(xs1)), rev(ys1)), col=gray(0.7))
polygon(c(xs2, rev(xs2)), c(rep(0, length(xs2)), rev(ys2)), col=gray(0.7))
#text(xs1[1], dt(0,18)/1.5, "t=-2.10\np=0.05", pos=3)
#text(xs2[1], dt(0,18)/1.5, "t=2.10\np=0.05", pos=3)
text(xs1[1], dt(0,18)/1.5, "critical value\nt=-2.10", pos=3)
text(xs2[1], dt(0,18)/1.5, "critical value\nt=2.10", pos=3)
arrows(xs1[1], dt(0,18)/1.5, xs1[1], ys1[1], length=0.1)
arrows(xs2[1], dt(0,18)/1.5, xs2[1], ys2[1], length=0.1)
#text(-3, dt(0,18), "t=-3\np=0.0038", pos=3)
#text(3, dt(0,18), "t=3\np=0.0038", pos=3)
#text(-3, dt(0,18), "t=-3", pos=3)
#text(3, dt(0,18), "t=3", pos=3)
#arrows(-3, dt(0,18), -3, dt(-3,18), length=0.1)
#arrows(3, dt(0,18), 3, dt(3,18), length=0.1)
par(op)
```

## Exercise 1

1) Open your `FANR6750` RStudio project (if you have one)  

2) Create a new `R` script and save it to the directory where you store you lab activities. Name it something like `lab02-t_tests.R`

3) Load the `FANR6750` package and the `treedata` object

```
library(FANR6750)
data("treedata)
```

4) Create 2 objects: `yL` is the tree density data for the first 10 experimental units (low elevation), and `yH` is the tree density data for the last 10 units (high elevation)

```
yL <- treedata$density[treedata$elevation == "Low"]

yH <- treedata$density[treedata$elevation == "High"]
```

5) Create objects that contain the mean, variance, and standard deviation of the 2 samples

```
mean.L <- ?
mean.H <- ?

s2.L <- ?
s2.H <- ?

s.L <- ?
s.H <- ?
```

# Graphics

`R` has very powerful graphing capabilities that make it possible to create data visualizations for reports or publications. As with most tasks in `R`, there are many ways to create graphs and you will find that people have very strong feelings about the best approach. 

The debate over graphics in `R` usually boils down to using the built-in graphing functions ("base graphics") vs the `ggplot2` package. Full disclosure, I much prefer `ggplot2` and therefore most of the sample code provided in lab will reflect that preference. However, **I don't care how you make your plots as long as they effectively display the information you are trying to convey**. If you prefer base graphics, by all means use base graphics.  

### Brief introduction to `ggplot2`

Because the code I provide will use `ggplot2`, it is worth briefly learning/reviewing how this package approaches data visualization. 

The power and flexibility of `ggplot2` come from it's consistent structure. Although a bit confusing at first, once you get the hang of it, the structure actually makes it quite easy to create highly customized visualizations. All plots created using `ggplot2` use the same underlying structure:

$$\underbrace{ggplot}_{initiate\; plot}(\underbrace{data = df}_{data\;frame},\; \underbrace{aes(x =\; , y = \;)}_{plot\; attributes}) + \underbrace{geom\_line()}_{geometry}$$

The `ggplot()` function initiates a new plot. In this function, you tell `ggplot2` what data frame you will be using for the plot (`ggplot` only accepts data frames as input) and you tell it how to map attributes of the data to the visual properties of the figures. Attributes are mapped inside the `aes()` argument. Attributes usually include location (`x-axis` and `y-axis` placement), color, size, shape, line type, and many others. In general, **each attribute will be mapped to one column of your data frame**. 

The `ggplot()` function simply initiates a graph - if you run just that portion of the code you will get a blank graph. We can see that by creating a new plot showing the relationship between `elevation` (the x-axis of the plot) and `density` (the y-axis): 

```{r fig.width=5, fig.height=4, fig.align='center'}
ggplot(data = treedata, aes(x = elevation, y = density))
```

You can see that `ggplot` created a figure with the correct axes and labels. But no data. That's because we didn't tell `ggplot` what type of *geometry* to use to represent the data. Geometry refers to the  type geometric object(s) we want to use to display the data. Common geometries include points (e.g., scatter plot), lines (e.g., time series), and bars (e.g., histograms). There are many others. Once we add a geometry, we can see the data:

```{r fig.width=5, fig.height=4, fig.align='center'}
ggplot(data = treedata, aes(x = elevation, y = density)) + 
  geom_point()
```

In this case, a boxplot might make more sense:

```{r fig.width=5, fig.height=4, fig.align='center'}
ggplot(data = treedata, aes(x = elevation, y = density)) + 
  geom_boxplot()
```

It's also possible to use more than one geometry:

```{r fig.width=5, fig.height=4, fig.align='center'}
ggplot(data = treedata, aes(x = elevation, y = density)) + 
  geom_boxplot() +
  geom_point()
```

This is reasonable figure showing tree densitys as a function of elevation. But `ggplot2` makes it very easy to tweak the way the data is visualized (maybe too easy, you can spend *a lot* of time tweaking minor details). For example, maybe we want to color the points based on the elevation. Because we want to map an *attribute* (color) to a *variable* (elevation), we make this change inside of `aes`:

```{r fig.width=5, fig.height=4, fig.align='center'}
ggplot(data = treedata, aes(x = elevation, y = density, color = elevation)) + 
  geom_boxplot() +
  geom_point()
```

That's not exactly what we wanted. Both the boxplot and the points now colored a function of `elevation`. To make just the points a function of elevation, we specify `color = elevation` inside of the `geom_point()` function (anything in the `ggplot()` function will apply to all geoms):

```{r fig.width=5, fig.height=4, fig.align='center'}
ggplot(data = treedata, aes(x = elevation, y = density)) + 
  geom_boxplot() +
  geom_point(aes(color = elevation))
```

We can also do things like the change the size of the geometries. In this case, we are **not** mapping a variable to an attribute (size is not a function of the data values). So these changes happen **outside** of the `aes()` argument:

```{r fig.width=5, fig.height=4, fig.align='center'}
ggplot(data = treedata, aes(x = elevation, y = density)) + 
  geom_boxplot() +
  geom_point(aes(color = elevation), size = 5)
```

One last example. Because many of the points overlap, it can be hard to tell how many individual points there are in each group. One way to deal with overplotting like this is to make each point slightly transparent. We can do that with the `alpha` parameter:

```{r fig.width=5, fig.height=4, fig.align='center'}
ggplot(data = treedata, aes(x = elevation, y = density)) + 
  geom_boxplot() +
  geom_point(aes(color = elevation), size = 5, alpha = 0.5)
```

Again, because we aren't mapping the alpha value to any data, we include it outside of `aes()`.

## Exercise 2

The graph above is fine for a quick visualization of the data but wouldn't be appropriate for including in publications or reports. On your own,

1) Improve the axis labels. This could include: title case, units, font size, etc. Run `?scale_y_continuous` and `?scale_x_discrete` if you need some help (and note the difference between these two functions!). `?theme` may also be useful

2) Manually change the color of the points (`?scale_color_manual`)

3) Instead of displaying the data using a boxplot, create histograms showing the distribution of tree densities at each elevation (`?geom_histogram`)

As you learn about graphics functions, whether base or `ggplot2`, you will probably need to look up help for how to do certain things. Google is usually your best bet but here are a few other good references:

- The fantastic [Fundamentals of Data Visualization](https://clauswilke.com/dataviz/) book by Claus Wilke

- The `ggplot2` package [website](https://ggplot2.tidyverse.org/)

- As, yes, even the base `R` [graph gallery](https://www.r-graph-gallery.com/base-R.html)

# t-tests

## Two-sample t-test with equal variances

Back to the t-test. Remember the basic steps for performing a two-sample t-test:

### Step 1: Compute the $t$ statistic:

$$\large t = \frac{(\bar{y}_L − \bar{y}_H) − (\mu_L − \mu_H)}{\sqrt{s^2_p/n_L + s^2_p/n_H}}$$
where $s^2_p$ is the pooled variance

$$\large s^2_p = \frac{(n_L − 1)s^2_L + (n_H − 1)s^2_H}{n_L + n_H − 2}$$

#### Step 2: Compare $t$ statistic to critical values:

- Critical values for 1-tailed test $t_{\alpha=0.05,18} = −1.73$ or $1.73$  

- Critical values for 2-tailed test $t_{\alpha=0.05,18} = −2.10$ and $2.10$

Now let's use `R` to run a t-test on the example data. 

## Option 1: Do it by hand

#### Step 1: Compute the $t$ statistic:

```{r echo = TRUE}
library(FANR6750)
data("treedata")

# Subset data
yL <- treedata$density[treedata$elevation == "Low"]
yH <- treedata$density[treedata$elevation == "High"]

# Compute means
mean.L <- mean(yL)
mean.H <- mean(yH)

# Compute variances
s2.L <- var(yL)
s2.H <- var(yH)

# Sample sizes
n.L <- length(yL)
n.H <- length(yH)

# Pooled variance
s2.p <- ((n.L-1)*s2.L + (n.H-1)*s2.H)/(n.L+n.H-2)

# Pooled standard error
SE <- sqrt(s2.p/n.L + s2.p/n.H)

# Compute t-statistic
t.stat <- (mean.L - mean.H) / SE
t.stat
```

#### Step 2: Compare $t$ statistic to critical values (two-tailed)

```{r echo = TRUE}
alpha <- 0.05

## NOTE: qt returns critical values from the t-distribution. No need to use "t tables"
critical.vals <- qt(c(alpha/2, 1-alpha/2), df=n.L+n.H-2)
critical.vals
```

## Option 2: Let R do all the work

```{r echo = TRUE}
t.test(yH, yL, var.equal = TRUE, paired = FALSE, alternative = "two.sided")
```

***
Make sure you set `var.equal=TRUE`. Otherwise, `R` will assume that the variances of the two populations are unequal  

***

## Option 1: Let R do all the work (alternate syntax)

```{r echo = TRUE}
t.test(density ~ elevation, data = treedata, 
       var.equal = TRUE, paired = FALSE, 
       alternative = "two.sided")
```

***
This second option returns identical results (though note that the order of the levels has been switched. Why is that?), but it is preferred because the notation is much more similar to the notation used to fit ANOVA models  

***

# Test equality of variances using `var.test`

```{r echo = TRUE}
var.test(yL, yH)
```

# Paired *t*-test

### Suppose the samples are paired

The caterpillar data:

```{r echo = TRUE}
data("caterpillar")
```

### For paired *t*-tests, we want to know if the mean of the differences differs from zero

```{r echo = TRUE}
caterpillar <- dplyr::mutate(caterpillar, diff = untreated - treated)
caterpillar$diff

mean(caterpillar$diff)
```

## Is the mean different from zero?

```{r echo = TRUE, fig.height = 5, fig.width=5, fig.density=3, fig.align='center'}
ggplot(data = caterpillar, aes(y = diff)) + 
  geom_boxplot() +
  geom_hline(yintercept = 0, linetype = "dashed")
```

## Paired *t*-test

**Recall**: Paired *t*-test is the same as a one-sample t-test on the differences. The hypothesis in the caterpillar example is one-tailed:

- $\large H_0 : \mu_d = 0$  

- $\large H_A : \mu_d > 0$  

#### Step 1. Calculate the standard deviation of the differences

$$\large s_d = \sqrt{\frac{1}{n-1}\sum_{i=1}^n(y_i - \bar{y})^2}$$

#### Step 2. Calculate the test statistic

$$\large t = \frac{\bar{y}-0}{s_d/\sqrt{n}}$$

#### Step 3. Compare to critical value

# Assignment

Create an R Markdown file to do the following:

1) Create an `R` chunk to load the caterpillar data using:

```{r eval = FALSE}
library(FANR6750)
data("caterpillar")
```

2) Make a figure to show the caterpillar counts on treated and untreated plots. If you plan to use `ggplot2`, you may want to learn about the `pivot_longer()` function from the `tidyr` package to reformat the data frame in the "long" format, which will make the plotting code easier.  

3) Create a header called "Hypotheses" and under this header, in plain English indicate what the null and alternative hypotheses are for the paired *t*-test. Also use R Markdown's equation features to write these hypotheses  using mathematical notation

4) Create a header called "Paired *t*-test by hand". Under this header, do a paired *t*-test on the caterpillar data without using the `t.test()` function. Use only the functions `mean`, `sd`, and possibly `length`. Be sure to **annotate your code** (either within the `R` chunk using `#` or as plain text within the R Markdown file) and **state the decision** (reject or accept the null) based on your results. 

5) Create a header called "Paired *t*-test in R". Under this header, do the paired *t*-test again, but this time using the `t.test` function.

    - You will need to use the paired argument when using the `t.test` function  

    - Assume variances are equal  

6) Create a header called "Unpaired *t*-test in R". Under this header, do a standard (unpaired) two-sample *t*-test using the `t.test()` function  

7) Add a section at the end of your file called "Discussion" and explain why the results differ between the paired and unpaired analysis  

A few things to remember when creating the assignment:

- Be sure the output type is set to: `output: html_document`

- Title the document: `title: "Lab 2 assignment"`

- Be sure to include your first and last name in the `author` section 

- Be sure to set `echo = TRUE` in all `R` chunks so we can see both your code and the output

- Regularly knit the document as you work to check for errors

- See the R Markdown reference sheet for help with creating `R` chunks, equations, tables, etc.


### Upload the `.Rmd` and  `.html` files to ELC before next week’s lab  

## Read pp. 127–131 in “Introductory Statistics with R”
