---
title: "Lab 8: Factorial designs"
subtitle: "FANR 6750: Experimental Methods in Forestry and Natural Resources Research"
author: "Fall 2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{lab07_factorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.align = 'center', fig.height = 4, fig.width = 4,
  fig.retina = 2,
  warning = FALSE, message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
source(here::here("R/zzz.R"))
library(kableExtra)
```

# Situation

When do we use a factorial design? Generally, when:

- There are 2 factors thought to influence the response variable

- The effect of each factor might depend on the other factor

- We have replicates for each combination of factors

# Example: effects of food and predators on *microtus* voles

## The data

```{r eval = FALSE}
library(FANR6750)
data("microtusData")

head(microtusData)
```

```{r echo = FALSE}
library(FANR6750)
data("microtusData")

kable(head(microtusData), align = "c")
```

```{r}
str(microtusData)
```

As before, we need to convert `food` to a factor:

```{r}
microtusData$food <- factor(microtusData$food)
str(microtusData)
```

How many replicates do we have?

```{r}
table(microtusData[,c("predators","food")])
```

## Visualizing the data

Visualizing multiple factors at once can sometimes be tricky. We'll use the `fill` aestethic to make separate boxplots for each predation level, within each food treatment:

```{r fig.width=6, fig.height=5}
ggplot(microtusData, aes(x = food, y = voles, fill = predators)) +
  geom_boxplot() +
  scale_x_discrete("Food treatment") +
  scale_y_continuous("Number of voles")
```

It's clear from this figure that 1) food supplementation influences vole abundance, and 2) the effects of food treatment depend on whether predators are present or absent. 

# Factorial design using `aov()`

We can analyze the *microtus* data using the `aov()` function that we're already familiar with. Note that to include the interaction of food and predators, we use `food * predators` in the formula:

```{r}
## food * predators --> food + predators + food:predators
aov1 <- aov(voles ~ food * predators, data = microtusData)
summary(aov1)
```

***

On your own, fit the model with the interaction (`food + predators`). How do the results compare?

***

## Follow ups

Often in your own analysis, you will not stop at a significant ANOVA result. For example, maybe you want to further explore the effects of one factor while holding the other factor constant. In this case, we can further test whether food supplements influence vole abundance *when predators are present* using the `subset` argument:

```{r}
summary(aov(voles ~ food, 
            data = microtusData, 
            subset = predators == "Present"))
```

Or when predators are absent:

```{r}
summary(aov(voles ~ food, 
            data = microtusData, 
            subset = predators == "Absent"))
```

Or maybe we want to test which levels are different using our old friend the Tukey HSD test, which for the factorial design will return both the "main" effect differences and the interaction:

```{r}
TukeyHSD(aov1)
```

# Visualizing the results

Another follow up that will often be important is visualizing the results of the ANOVA model. In the case of the factorial design, this will generally involve making a graph of the predicted response (and associated uncertainty!) for each combination of factors. 

To create the data that will form the basis of this visualization, we'll first use the `model.tables()` function to compute the relevant means (i.e., predicted values) and standard errors:

```{r}
(tab <- model.tables(aov1, type="means", se = TRUE))
```

Next, we'll extract the group means

```{r}
(ybar_ij. <- tab$tables$"food:predators")
```

What about confidence intervals? Notice that `model.tables()` returns "Standard errors for
differences of means".  

Extending the formulas to the $A \times B$ factorial case, the confidence interval for the difference in two $A \times B$ means is:

$$\Large CI_{1−\alpha} : \bar{y}_{ij.} − \bar{y}_{ij.′} \pm t_{1−\alpha/2,ab(n−1)}\times \sqrt{\frac{2MSE}{n}}$$

where $n = 4$ in the vole example.

To set up a plot of the group means, however, we need a confidence interval for each mean. We take out the ‘2’ from the SE expression:

$$\Large CI_{1−\alpha} : \bar{y}_{ij.} \pm t_{1−\alpha/2,ab(n−1)}\times \sqrt{\frac{MSE}{n}}$$

Where can we find MSE?

Answer: in the output of `summary()` of our `aov1` object:

```{r}
str(summary(aov1))
```

Calculate the width of the confidence intervals:

```{r}
MSE <- summary(aov1)[[1]]$`Mean Sq`[4] ## 16.5
ybar_ij.SE <- sqrt(MSE/4)
CI.half <- qt(0.975, 18) * ybar_ij.SE
(CI <- c(-CI.half, CI.half))
```

Now we can create a data frame and plot the group means and confidence intervals:

```{r fig.width=6, fig.height=5}
predicted.voles <- data.frame(Food = rep(c(0, 1, 2), 2),
                              Predators = rep(c("Absent", "Present"), each = 3),
                              voles = c(ybar_ij.))

## Add lower and upper confidence intervals
predicted.voles <- dplyr::mutate(predicted.voles, 
                                 LCI = voles + CI[1],
                                 UCI = voles + CI[2])

## Plot results
ggplot(predicted.voles, aes(x = Food, y = voles, 
                            color = Predators)) +
  geom_path(aes(linetype = Predators)) +
  geom_errorbar(aes(ymin = LCI, ymax = UCI), width = 0) +
  geom_point() +
  scale_y_continuous("Predicted number of voles")
```


# Assignment (not for a grade)

**Fictitious Scenario**   

Acid rain has lowered the pH of many lakes in the northeastern United States, and as a result, fish populations have declined. Managers have resorted to aerial applications of lime (powdered calcium carbonate) in hopes of increasing pH.  

To determine if lime applications result in increased pH, they applied equal amounts of lime to 15 lakes, and as a control, they applied the same amount of inert white powder to an additional 15 lakes.  

Researchers suspected that the effect of lime might depend upon the buffering effects of the underlying bedrock.  

To assess this hypothesis, the 30 lakes were chosen such that 10 had limestone bedrock, 10 had granite bedrock, and 10 had schist bedrock. Acidity was measured before and after each application, and the difference in pH is recorded in the `acidityData` data set:

```{r}
data("acidityData")
head(acidityData)
```

**Questions/tasks**  

1) What are the null and alternative hypotheses?

2) Test the null hypotheses using an $A \times B$ factorial ANOVA implemented with `aov()`. 

3) Does the effect of lime depend upon the bedrock type? If so, how? 

Put your answers in an R Markdown report. Your report should include:

1) A well-formatted ANOVA table (with caption); and

2) A publication-quality plot of the estimated effect of lime on pH change, including 95% confidence intervals. The figure should also have a descriptive caption and any aesthetics (color, line type, etc.) should be clearly defined.

You may format the report however you like but it should be well-organized, with relevant headers, plain text, and the elements described above.  

As always:

- Be sure the output type is set to: `output: html_document`

- Title the document: `title: "Lab 7 assignment"`

- Be sure to include your first and last name in the `author` section 

- Be sure to set `echo = TRUE` in all `R` chunks so we can see both your code and the output

- See the R Markdown reference sheet for help with creating `R` chunks, equations, tables, etc.

- See the "Creating publication-quality graphics" reference sheet for tips on formatting figures





































