---
title: "Lab 10: Repeated measures"
subtitle: "FANR 6750: Experimental Methods in Forestry and Natural Resources Research"
author: "Fall 2021"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{lab10_repeated_measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.align = "center", fig.height = 5, fig.width = 6,
  fig.retina = 2,
  warning = FALSE, message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
source(here::here("R/zzz.R"))
library(kableExtra)
```

# Overview

Experiments that use "repeated measures" methods are those that have the following design:

- We randomly assign each ‘subject’ to a treatment  

- We record the response to the treatment over time

In these cases, we need to think about the following sources of variation:

- Treatment  

- Time

- Treatment-time interaction

- Random variation among subjects

- Random variation within subjects

## Approaches

As we saw in lecture, there are several ways to analyze repeated measures data, including:

#### Univariate

- This is similar to a split-plot (or nested and crossed) analysis, but with adjusted *p*-values  

- Adjustments assuming "near-sphericity": Greenhouse-Geisser or Huynh-Feldt methods  

- In `R`, you must do a MANOVA to obtain these adjusted *p*-values  

#### MANOVA

- Testing based on Wilks’ lambda or Pillai’s trace

- No assumptions about "sphericity" are required

- This is usually followed by a profile analysis

#### Mixed effects model with (ARMA) correlation structure

- This can be done using `lme()`

- We might cover this later

Let's see how to implement these approaches in `R`

## Univariate Split-plot Approach

First, the data:

```{r}
library(FANR6750)
data("plantData")
plantData$plant <- factor(plantData$plant)
plantData$fertilizer <- factor(plantData$fertilizer)
plantData$week <- factor(plantData$week)
str(plantData)
head(plantData, n = 8)
```

As we can see, these data contain information on leaf growth by 10 plants subjected to different fertilizer treatments. Leaf growth on each plant was measured weekly for 5 weeks, leading to the repeated measures structure. 

As always, it helps to visualize the data:

```{r fig.align="center"}
library(ggplot2)

ggplot(plantData, aes(x = week, y = leaves, 
                      group = plant, 
                      color = fertilizer)) +
  geom_path() +
  scale_x_discrete("Week") +
  scale_y_continuous("Number of leaves")
```

Now let's fit the univariate model:

```{r}
aov1 <- aov(leaves ~ fertilizer * week + Error(plant),
            data = plantData)
summary(aov1)
```

Although this output looks ok, remember that we need to adjust the *p*-values for the time and interaction effects. In `R`, this requires reformatting the data and running a MANOVA

Specifically, we need to covert the data from the "long" format that it is currently in (each measurement on each plant is contained in its own row) to the "wide" format (each plant is one row, with separate columns for each weekly measurement). There are a number of ways to convert between long and wide formats in `R` (all of which are, in my opinion, confusing and require looking up syntax/examples every time) but we'll use the `pivot_wider()` function from the `tidyr` package^[to run this function, we need to specify which column in the original data frame contains the column names (`names_from`) and which column contains the values that will go in the new columns (`values_from`). We can also (optionally) specify a prefix to add to each column name so that, as would the base here, we don't end up with numeric column names. For more help, see `?pivot_wider`]:

```{r}
plantData_wide <- tidyr::pivot_wider(data = plantData, 
                                     names_from = week, 
                                     names_prefix = "week",
                                     values_from = leaves)
head(plantData_wide)
```

Now we can run the MANOVA:

```{r}
manova1 <- manova(cbind(week1, week2, week3, week4, week5) ~ fertilizer, 
                  data = plantData_wide)
```

And the results will contain the adjusted *p*-values:

```{r}
anova(manova1, X = ~1, test = "Spherical")
```

### An aside – sphericity

*Sphericity* is the multivariate analogue of the homogeneity of variance assumption of ANOVA.

Technically, adjusted *p*-values and MANOVA aren’t necessary if the assumption of sphericity holds. However, Mauchly’s sphericity test has low power, and it is often recommended that the *p*-values be adjusted even if the test fails to reject the sphericity assumption.

```{r}
mauchly.test(manova1, X=~1)
```

In this case, we fail to reject the null hypothesis, so sphericity can be assumed, but the adjusted *p*-values should be reported anyway.

***

On your own, go back and compare the results from `aov()` without adjusting the *p*-values to the `anova()` results from the MANOVA model. How did they change? Did our decision about the effects of fertilizer on leaf growth change?

***

## Multivariate tests – Wilks’ lambda ($\Lambda$)

An alternative to the adjusted *p*-value approach is to do a multivariate analysis relaxing the assumptions about the structure of the variance-covariance matrix.

The good news is we’re already most of the way there:

```{r}
anova(manova1, X=~1, test="Wilks")
```

***

How do our conclusions from this approach compare to conclusions from the uni-variate approach?

***


Pillai’s trace is an alternative to Wilks’ lambda. In this case, it returns the same *p*-values as Wilks’ test:

```{r}
anova(manova1, X=~1, test="Pillai")
```

### Profile analysis – when do growth rates differ?

Profile analysis requires calculating the differences (i.e., the number of leaves grown each week):

```{r}
manova2 <- manova(cbind(week2 - week1, week3 - week2, week4 - week3, week5 - week4) ~ fertilizer, data = plantData_wide)
summary.aov(manova2)
```

We can also plot the growth rates. First, calculate mean growth rate for each time interval:

```{r}
leavesMat <- plantData_wide[,3:7]
growthMat <- leavesMat[,2:5] - leavesMat[,1:4]
colnames(growthMat) <- paste("interval", 1:4, sep=".")
(lowFertilizer <- colMeans(growthMat[1:5,]))

(highFertilizer <- colMeans(growthMat[6:10,]))
```

Next, calculate the standard errors for these growth rates:

```{r}
SE <- sqrt(diag(stats:::vcov.mlm(manova2)))
SE <- SE[names(SE)==":(Intercept)"] # Only use "intercept" SEs
unname(SE) ## Ignore the names
```

Now, create a data frame and plot the results:

```{r}
growthDF  <- data.frame(interval = rep(1:4, 2),
                        fertilizer = rep(c("Low", "High"), each = 4),
                        growth = c(lowFertilizer, highFertilizer), 
                        SE = rep(SE, 2))


ggplot(growthDF, aes(x = interval, y = growth, color = fertilizer)) +
  geom_path() +
  geom_point() +
  geom_errorbar(aes(ymin = growth - SE, ymax = growth + SE), width = 0.1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_x_continuous("Time interval") +
  scale_y_continuous("Growth rate (leaves/week)", limits  = c(-1, 4))
```

# Assignment

A researcher wants to assess the effects of crowding on the growth of the dark toadfish (*Neophrynichthys latus*). Fifteen fish tanks are stocked with three densities of toadfish. Five tanks have low density (1 fish), 5 tanks have medium density (5 fish), and 5 tanks have high density (10 fish). In each tank, the weight of one “focal fish” is recorded on 6 consecutive weeks. 
The data can be loaded using:

```{r}
library(FANR6750)
data("fishData")
```

1) Conduct the univariate repeated measures ANOVA using `aov()`. Calculate the adjusted *p*-values using the Huynh-Feldt method. Does the effect of density on growth change over time?

2) Conduct a multivariate repeated measures ANOVA and use Wilks’ lambda to test if the effect of density changes over time. What is your conclusion?

3) Conduct a profile analysis. In which time intervals is the effect of density
on growth rate significant?

Put your answers in an R Markdown report. Your report should include:

1) Well-formatted ANOVA tables (with captions); and

2) Publication-quality plots displaying the relationships being test and/or estimated effects. The figures should also have descriptive captions and any aesthetics (color, line type, etc.) should be clearly defined.

You may format the report however you like but it should be well-organized, with relevant headers, plain text, and the elements described above. 

As always:

- Be sure the output type is set to: `output: html_document`

- Title the document: `title: "Lab 10 assignment"`

- Be sure to include your first and last name in the `author` section 

- Be sure to set `echo = TRUE` in all `R` chunks so we can see both your code and the output

- Please upload both the `html` and `.Rmd` files when you submit your assignment  

- See the R Markdown reference sheet for help with creating `R` chunks, equations, tables, etc.



