---
title: "Lab 13: Generalized linear models"
subtitle: "FANR 6750: Experimental Methods in Forestry and Natural Resources Research"
author: "Fall 2021"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{lab13_glm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.align = 'center', fig.height = 4, fig.width = 6,
  fig.retina = 2,
  warning = FALSE, message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
source(here::here("R/zzz.R"))
library(kableExtra)
```

# Logistic regression

Remember from lecture the logistic regression model:

$$\Large logit(p_i) = \beta_0 + \beta_1x_{i1} + \beta_2x_{i2} + · · ·$$

$$\Large y_i \sim binomial(N, p_i)$$
where:

- $N$ is the number of ‘trials’ (e.g. coin flips)  

- $p_i$ is the probability of a success for sample unit $i$

## Example: Frog data

```{r eval = FALSE}
library(FANR6750)
data("frogData")
head(frogData)
```

```{r echo = FALSE}
library(FANR6750)
data("frogData")
kable(head(frogData))
```

Let's start by using the `glm()` function to model the relationship between frog detections (0 = not detected, 1 = detected), habitat, and elevation:

```{r}
fm1 <- glm(presence ~ habitat + elevation, family = binomial(link = "logit"), data = frogData)
summary(fm1)
```

Notice that we use the `family =` argument to tell `R` that this is a logistic (i.e., binomial) glm with a logit link function. 

***

What do the parameter estimates from this model represent, including the intercept? What can we say about the effects of habitat and elevation on the probability of detecting a frog? 

Use the `model.matrix()` function to examine the dummy variable structure and contrasts.

***

As we have seen previously, we can use `predict()` to estimate occurrence probability at different combinations of elevation and habitat. For example, how does probability of occurrence differ across elevations in oak habitat:

```{r eval = FALSE}
predData.elev <- data.frame(elevation = seq(12, 489, length = 50),habitat = "Oak")
head(predData.elev)
```

```{r echo = FALSE}
predData.elev <- data.frame(elevation = seq(12, 489, length = 50),habitat = "Oak")
kable(head(predData.elev))
```

By default, the predictions are on the link scale so to get confidence intervals on the probability scale, we have to back transform using the inverse-link (`plogis()`):

```{r}
pred.link <- predict(fm1, newdata = predData.elev, se.fit = TRUE)
predData.elev$p <- plogis(pred.link$fit) # back transform to probability scale
predData.elev$lower <- plogis(pred.link$fit - 1.96 * pred.link$se.fit)
predData.elev$upper <- plogis(pred.link$fit + 1.96 * pred.link$se.fit)
```

And now plot the predictions and observations:

```{r}
library(ggplot2)

ggplot() +
  geom_point(data = frogData, aes(x = elevation, y = presence)) +
  geom_path(data = predData.elev, aes(x = elevation, y = p)) +
  geom_ribbon(data = predData.elev, aes(x = elevation, ymin = lower, ymax = upper),
              fill = NA, color = "black", linetype = "longdash") +
  scale_y_continuous("Probability of occurrence") +
  scale_x_continuous("Elevation (m)")
```

What about occurrence probability in each habitat type:

```{r}
predData.hab <- data.frame(habitat = c("Oak", "Maple", "Pine"), elevation = 250)
pred.hab <- predict(fm1, newdata = predData.hab, se.fit = TRUE)
predData.hab$p <- plogis(pred.hab$fit) # back transform to probability scale
predData.hab$lower <- plogis(pred.hab$fit - pred.hab$se.fit)
predData.hab$upper <- plogis(pred.hab$fit + pred.hab$se.fit)

ggplot() +
  geom_col(data = predData.hab, aes(x = habitat, y = p), fill = "grey60") +
  geom_errorbar(data = predData.hab, aes(x = habitat, ymin = lower, ymax = upper),
                width = 0.1) +
  scale_y_continuous("Probability of occurrence") +
  scale_x_discrete("Habitat type")
```

# Poisson regression

Remember from lecture that the model for the Poisson regression is:

$$\Large log(\lambda_i) = \beta_0 + \beta_1x_{i1} + \beta_2x_{i2} + · · ·$$
$$\Large y_i \sim Poisson(\lambda_i)$$

where:

- $\lambda_i$ is the expected value of $y_i$

***

In the linear regression model we saw in previous weeks, we had a variance term $sigma^2$ that measured on far, on average, each response was from it's expected value. But neither of the two generalized linear models we learned about here have an equivalent variance term. Why is that?

***

Using the frog data, let's model the number of frogs detected at each point as a function of habitat and elevation:

```{r}
fm2 <- glm(abundance ~ habitat + elevation, family = poisson(link = "log"), data = frogData)
summary(fm2)
```

Again, be sure you can interpret the parameters in this model, including the intercept. 

To get confidence intervals on the $(0, \infty)$ scale, predict on the log (link) scale and then back transform using the inverse-link (`exp()`):

```{r}
pred.link <- predict(fm2, newdata = predData.elev, se.fit = TRUE)
predData.elev$lambda <- exp(pred.link$fit) # exp is the inverse-link function
predData.elev$lower <- exp(pred.link$fit - 1.96 * pred.link$se.fit)
predData.elev$upper <- exp(pred.link$fit + 1.96 * pred.link$se.fit)

ggplot() +
  geom_point(data = frogData, aes(x = elevation, y = abundance)) +
  geom_path(data = predData.elev, aes(x = elevation, y = lambda)) +
  geom_ribbon(data = predData.elev, aes(x = elevation, ymin = lower, ymax = upper),
              fill = NA, color = "black", linetype = "longdash") +
  scale_x_continuous("Elevation") +
  scale_y_continuous("Expected abundance")
```

# Assignment

Researchers want to know how latitude and landscape type influence the probability that American Crows are infected by West Nile Virus. One hundred crows are captured and tested for West Nile Virus in urban and rural landscapes spanning a latitude gradient.

The data can be loaded using:

```{r}
library(FANR6750)
data("crowData")
```

Create an R markdown report to do the following:

1) Fit a logistic regression model to the crow data to assess the effects of latitude and landscape type

2) Include a well-formatted summary table of the parameter estimates

3) Provide a clear interpretation of the parameter estimates in the text

4) Plot the relationship between infection probability and latitude, for rural and urban landscapes, on the same graph. The graph should include:

    + the observed data points (color coded by landscape)  
    
    + a legend  
    
    + confidence intervals  

You may format the report however you like but it should be well-organized, with relevant headers, plain text, and the elements described above. 

As always:

- Be sure the output type is set to: `output: html_document`

- Title the document: `title: "Lab 13 assignment"`

- Be sure to include your first and last name in the `author` section 

- Be sure to set `echo = TRUE` in all `R` chunks so we can see both your code and the output

- Please upload both the `html` and `.Rmd` files when you submit your assignment  

- See the R Markdown reference sheet for help with creating `R` chunks, equations, tables, etc.
