---
title: "LECTURE 18: generalized linear models"
subtitle: "FANR 6750 (Experimental design)"
author: "<br/><br/><br/>Fall 2021"
output:
  xaringan::moon_reader:
    css: ["default", "FANR6750.css", "FANR6750-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, echo = FALSE, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning=FALSE, message=FALSE, 
                      fig.retina = 2, fig.width = 9, fig.height = 6)
source(here::here("R/zzz.R"))
library(emo)
library(kableExtra)
library(dplyr)
library(FANR6750)
# library(gganimate)
```

# outline 

<br/>

1) Motivation

<br/>

--

2) Logistic regression

<br/>

--

3) Poisson regression



---
# motivation

#### Benefits of generalized linear models  

--

- The residuals don’t have to be normally distributed  

--

- The response variable can be binary, integer, strictly-positive, etc...  

--

- The variance is not assumed to be constant  

--

- Useful for manipulative experiments or observational studies, just like linear models 

--

#### Examples

- Presence-absence studies  

- Studies of survival  

- Seed germination studies  

---
# from linear to generalized linear

#### Linear model

$$\Large E(y_i) = \beta_0 + \beta_1x_{i1} +\beta_2x_{i2} + ...$$
$$\Large y_i \sim normal(E(y_i), \sigma^2)$$

--

#### Generalized linear model

$$\Large g\big(E(y_i)\big) = \beta_0 + \beta_1x_{i1} +\beta_2x_{i2} + ...$$
$$\Large y_i \sim f(E(y_i))$$

--

where $g()$ is a **link function**, such as the log or logit link and $f$ is a probability distribution such as the binomial or Poisson

---
# alternative representations

#### This:
$$\Large g\big(E(y_i)\big) = \beta_0 + \beta_1x_{i1} +\beta_2x_{i2} + ...$$
$$\Large y_i \sim f(E(y_i))$$

#### Is the same as this:

$$\Large E(y_i) = g^{-1}(\beta_0 + \beta_1x_{i1} +\beta_2x_{i2} + ...)$$
$$\Large y_i \sim f(E(y_i))$$

--

#### Is the same as this:

$$\Large g\big(E(y_i)\big) = \mathbf{X \beta}$$
$$\Large y_i \sim f(E(y_i))$$

---
# link functions

```{r}
link <- data.frame(Distribution = c("Binomial", "Poisson"),
                   link = c("logit", "log"),
                   equation = c('\\(log(\\frac{p}{1-p})\\)',
                                '\\(log(\\lambda)\\)'),
                   inverse = c('\\(\\frac{e^{X\\beta}}{1+e^{X\\beta}}\\)',
                                '\\(e^{X\\beta}\\)'))

link %>%
  kable(format = 'html', col.names = c("Distribution", "link name", "link equation", "inverse link equation"), escape = FALSE, align = "c")
```

<br/>

--

```{r}
linkR <- data.frame(Distribution = c("Binomial", "Poisson"),
                   link = c("logit", "log"),
                   equation = c("qlogis", "log"),
                   inverse = c("plogis", "exp"))

linkR %>%
  kable(format = 'html', col.names = c("Distribution", "link name", "link in R", "inverse link in R"), align = "c")
```

---
# logit link example

```{r echo = TRUE}
beta0 <- 5
beta1 <- -0.08
(logit.p <- beta0 + beta1*100) # linear predictor for x = 100
```

--

#### How do we convert -3 to a probability? Use the inverse-link:

```{r echo = TRUE}
(p <- exp(logit.p)/(1+exp(logit.p))) # same as plogis(logit.p)
```

--

#### To go back, use the link function itself:

```{r echo =TRUE}
log(p/(1-p))
qlogis(p)
```

---
# logit link example

```{r echo = TRUE}
plot(function(x) 5 + -0.08*x, from=0, to=100,
xlab="Elevation", ylab="logit(prob of occurrence)")
```

---
# logit link example

```{r echo = TRUE}
plot(function(x) plogis(5 + -0.08*x), from=0, to=100,
xlab="Elevation", ylab="Probability of occurrence")
```

---
class: inverse, middle, center
# logistic regression

---
# logistic regression

#### Logistic regression is a specific type of GLM in which the response variable follows a binomial distribution and the link function is the logit  

--

#### It would be better to call it “binomial regression” since other link functions (e.g. the probit) can be used  

--

#### Appropriate when the response is binary or a count with an upper limit

--

#### Examples:

- Presence/absence studies  

- Survival studies  

- Disease prevalance studies  


---
# logistic regression

<br/>

$$\Large logit(p_i) = \beta_0 + \beta_1x_{i1} + \beta_2x_{i2} + ...$$
$$\Large y_i \sim binomial(N, p_i)$$
--

#### where:

- $\large N$ is the number of “trials” (e.g. coin flips)  

- $\large p_i$ is the probability of success for sample unit *i*

---
# binomial distribution

<br/>

```{r}
bindist <- data.frame(y1 = rbinom(100, 5, 0.5),
                      y2 = rbinom(100, 5, 0.9))

ggplot(bindist, aes(x = y1)) +
  geom_histogram(aes(y = stat(count) / sum(count)), 
                 binwidth = 1, fill = "grey70", color = "black", ) +
  scale_y_continuous("Probability") +
  scale_x_continuous("Number of successes", breaks = 0:5, limits = c(-0.5, 5.5)) +
  labs(title = "Binomial(N = 5, p = 0.5)")
```

---
# binomial distribution

<br/>

```{r}
ggplot(bindist, aes(x = y2)) +
  geom_histogram(aes(y = stat(count) / sum(count)), 
                 binwidth = 1, fill = "grey70", color = "black", ) +
  scale_y_continuous("Probability") +
  scale_x_continuous("Number of successes", breaks = 0:5, limits = c(-0.5, 5.5)) +
  labs(title = "Binomial(N = 5, p = 0.9)")
```

---
# binomial distribution

#### Properties

- The expected value of $\large y$ is $\large Np$  

- The variance is $\large Np(1 - p)$  

--

#### Bernoulli distribution

- The Bernoulli distribution is a binomial distribution with a single trial (N = 1)  

- Logistic regression is usually done in this context, such that the response variable is 0/1 or No/Yes or Bad/Good, etc.

---
# worked example using glm

First we will model the presence-absence response variable to determine if elevation and habitat affect the probability of occurrence

```{r echo = TRUE}
data("frogData")
head(frogData, n=12)
```

---
# raw data

```{r echo = TRUE}
ggplot(frogData, aes(x = elevation, y = presence)) +
  geom_point() +
  scale_y_continuous("Frog Occurrence") + scale_x_continuous("Elevation")
```

---
# raw data

```{r echo = TRUE, fig.height=5, fig.width=8}
library(dplyr)

frogData %>% group_by(habitat) %>% summarise(group.prob = mean(presence)) %>%
  ggplot(., aes(x = habitat, y = group.prob)) +
  geom_col(fill = "grey70", color = "black") +
  scale_y_continuous("Proportion of sites with frogs") + scale_x_discrete("Habitat")
```

---
# the `glm` function

<br/>

```{r echo = TRUE, eval = FALSE}
fm1 <- glm(presence ~ habitat + elevation, 
           family=binomial(link="logit"), 
           data = frogData)
broom::tidy(fm1)
```

<br/>

```{r echo = FALSE, eval = TRUE}
fm1 <- glm(presence ~ habitat + elevation, family=binomial(link="logit"), data = frogData)
kable(broom::tidy(fm1))
```

---
# occurrence probability and elevation

```{r echo = TRUE}
newdat <- data.frame(elevation=seq(12, 489, length = 50), 
                     habitat="Oak")
head(newdat)
```

--

To get confidence intervals on (0,1) scale, predict on linear (link) scale and then backtransform using inverse-link

```{r echo = TRUE}
pred.link <- predict(fm1, newdata=newdat, se.fit=TRUE, type="link")
newdat$mu <- plogis(pred.link$fit)
newdat$lower <- plogis(pred.link$fit - 1.96*pred.link$se.fit)
newdat$upper <- plogis(pred.link$fit + 1.96*pred.link$se.fit)
```

---
# occurrence probability and elevation

```{r}
ggplot(frogData, aes(x = elevation, y = presence)) +
  geom_point() +
  geom_line(data = newdat, aes(x = elevation, y = mu)) +
  geom_line(data = newdat, aes(x = elevation, y = lower), linetype = "longdash") +
  geom_line(data = newdat, aes(x = elevation, y = upper), linetype = "longdash") +
  scale_y_continuous("Probability of Occurrence") + scale_x_continuous("Elevation")

```

---
class: center, middle, inverse

# poisson regression

---
# poisson regression

<br/>

$$\Large log(\lambda_i) = \beta_0 + \beta_1x_{i1} + \beta_2x_{i2} + ...$$
$$\Large y_i \sim Poisson(\lambda_i)$$

<br/>

--

#### where

- $\large \lambda_i$ is the expected value of $\large y_i$

---
# poisson regression

#### Useful for:

- Count data  

- Number of events in time intervals  

- Other types of integer data  

--

#### Properties

- The expected value of *y* ( $\large \lambda$) is equal to the variance  

- This is an assumption of the Poisson model  

- Like all assumptions, it can be relaxed if you have enough data  

---
# poisson distribution

<br/>

```{r}
poisdist <- data.frame(y1 = rpois(500, 1),
                       y2 = rpois(500, 5),
                       y3 = rpois(500, 10))

ggplot(poisdist, aes(x = y1)) +
  geom_histogram(aes(y = stat(count) / sum(count)), 
                 binwidth = 1, fill = "grey70", color = "black", ) +
  scale_y_continuous("Probability") +
  scale_x_continuous("Reponse variable", breaks = c(0, 5,10,15,20,25), limits = c(-0.5, 25.5)) +
  labs(title = "Poisson(1)")
```

---
# poisson distribution

<br/>

```{r}
ggplot(poisdist, aes(x = y2)) +
  geom_histogram(aes(y = stat(count) / sum(count)), 
                 binwidth = 1, fill = "grey70", color = "black", ) +
  scale_y_continuous("Probability") +
  scale_x_continuous("Reponse variable", breaks = c(0, 5,10,15,20,25), limits = c(-0.5, 25.5)) +
  labs(title = "Poisson(5)")
```

---
# poisson distribution

<br/>

```{r}
ggplot(poisdist, aes(x = y3)) +
  geom_histogram(aes(y = stat(count) / sum(count)), 
                 binwidth = 1, fill = "grey70", color = "black", ) +
  scale_y_continuous("Probability") +
  scale_x_continuous("Reponse variable", breaks = c(0, 5,10,15,20,25), limits = c(-0.5, 25.5)) +
  labs(title = "Poisson(10)")
```

---
# log link example

```{r echo = TRUE}
plot(function(x) 5 + -0.08*x, from=0, to=100,
xlab="Elevation", ylab="log(Expected abundance)")
```

---
# log link example

```{r echo = TRUE}
plot(function(x) exp(5 + -0.08*x), from=0, to=100,
xlab="Elevation", ylab="Expected abundance")
```

---
# the `glm` function

<br/>

```{r echo = TRUE, eval = FALSE}
fm2 <- glm(abundance ~ habitat + elevation, family=poisson(link="log"), 
           data=frogData)
broom::tidy(fm2)
```

<br/>

```{r echo = FALSE, eval = TRUE}
fm2 <- glm(abundance ~ habitat + elevation, family=poisson(link="log"), 
           data=frogData)
kable(broom::tidy(fm2))
```

---
# prediction

```{r echo = TRUE}
newdat <- data.frame(elevation = seq(12, 489, length = 50),
                     habitat = "Oak")
head(newdat)
```

--

To get confidence intervals on (0, $\infty$) scale, predict on linear (link) scale and then backtransform using inverse-link

```{r echo = TRUE}
pred.link <- predict(fm2, newdata=newdat, se.fit=TRUE, type="link")
newdat$mu <- exp(pred.link$fit)
newdat$lower <- exp(pred.link$fit - 1.96*pred.link$se.fit)
newdat$upper <- exp(pred.link$fit + 1.96*pred.link$se.fit)
```

---
# prediction

<br/>

```{r}
ggplot(frogData, aes(x = elevation, y = abundance)) +
  geom_point() +
  geom_line(data = newdat, aes(x = elevation, y = mu)) +
  geom_line(data = newdat, aes(x = elevation, y = lower), linetype = "longdash") +
  geom_line(data = newdat, aes(x = elevation, y = upper), linetype = "longdash") +
  scale_y_continuous("Probability of Occurrence") + scale_x_continuous("Elevation")

```

---
# assessing model fit

#### The most common problem in Poisson regression is overdispersion  

--

#### Overdispersion is the situation in which there is more variability in the data than predicted by the model  

--

#### Overdispersion cannot be assessed by simply comparing the mean and variance of the response variable  

--

### The presence of many zeros is not necessarily indicative of overdispersion  

--

### Overdispersion can be assessed using a goodness-of-fit test  

---
# goodness-of-fit

#### The fit of a Poisson regression can be assessed using a $\large \chi^2$ test  

The test statistic is the residual deviance:

$$\Large D = 2\bigg[\sum y_i log\bigg(\frac{y_i}{\hat{\lambda}_i}\bigg)-(y_i-\hat{\lambda}_i)\bigg]$$

--

If the null hypothesis is true (ie, the model fits the data), $D$ should follow $\chi^2$ distribution with $N - K$ degrees-of-freedom  

--

```{r echo = TRUE}
N <- nrow(frogData) # sample size
K <- length(coef(fm2)) # number of parameters
df.resid <- N-K # degrees-of-freedom
Dev <- deviance(fm2) # residual deviance
(p.value <- 1-pchisq(Dev, df=df.resid))# p-value
```

---
### $\large \chi^2$ DISTRIBUTION AND RESIDUAL DEVIANCE

<br/>

```{r echo = TRUE, fig.height=5, fig.width=7}
curve(dchisq(x, df=df.resid), from=0, to=50, xlab="Deviance", ylab="Density")
abline(v=Dev, lwd=3, col="red")
```

--
The red line is the residual deviance. We reject the null hypothesis, and we conclude that the Poisson model does not fit the data

---
# what if model doesn't fit the data?

<br/>

#### Alternatives to the Poisson distribution

- Negative binomial  

- Zero-inflated Poisson

---
# negative binomial distribution

```{r}
nb_df <- data.frame(y1 = rnbinom(500, mu = 2, size = 10),
                    y2 = rnbinom(500, mu = 2, size = 5),
                    y3 = rnbinom(500, mu = 2, size = 0.1))

ggplot(nb_df, aes(x = y1)) +
  geom_histogram(aes(y = stat(count) / sum(count)), 
                 binwidth = 1, fill = "grey70", color = "black", ) +
  scale_y_continuous("Probability") +
  scale_x_continuous("Reponse variable", breaks = c(0, 5,10,15,20,25), limits = c(-0.5, 25.5)) +
  labs(title = "NegBin(2,10)")

```

---
# negative binomial distribution

```{r}

ggplot(nb_df, aes(x = y2)) +
  geom_histogram(aes(y = stat(count) / sum(count)), 
                 binwidth = 1, fill = "grey70", color = "black", ) +
  scale_y_continuous("Probability") +
  scale_x_continuous("Reponse variable", breaks = c(0, 5,10,15,20,25), limits = c(-0.5, 25.5)) +
  labs(title = "NegBin(2, 5)")

```

---
# negative binomial distribution

```{r}

ggplot(nb_df, aes(x = y3)) +
  geom_histogram(aes(y = stat(count) / sum(count)), 
                 binwidth = 1, fill = "grey70", color = "black", ) +
  scale_y_continuous("Probability") +
  scale_x_continuous("Reponse variable", breaks = c(0, 5,10,15,20,25), limits = c(-0.5, 25.5)) +
  labs(title = "NegBin(2, 0.1)")

```

