---
title: "LECTURE 11: fixed effects and random effects"
subtitle: "FANR 6750 (Experimental design)"
author: "<br/><br/><br/>Fall 2021"
output:
  xaringan::moon_reader:
    css: ["default", "FANR6750.css", "FANR6750-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, echo = FALSE, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning=FALSE, message=FALSE, 
                      fig.retina = 2, fig.width = 9, fig.height = 7)
library(WILD3810)
library(emo)
library(kableExtra)
library(dplyr)
library(FANR6750)
# library(gganimate)
```

# outline 

<br/>

1) Motivation

<br/>

--

2) Random effects in one-way ANOVA

<br/>

--

3) Random effects in blocked ANOVA


---
# motivation

#### The completely randomized ANOVA we have discussed can be applied to many different experiments. For example, it could be used to:  

- Assess the best option from two different pesticides (plus no spray) used for gypsy moth control  

- Assess if there is significant variability among the test scores of introductory statistics classes when taught by different instructors  

--

#### In the first example, we would model the effect of each pesticide as a fixed effect  

--

#### In the second example, the effect of each instructor would be modeled as a random effect  

---
# fixed vs random effects

<br/>

#### A fixed effects model is appropriate when the treatment levels included in the experiment are exhaustive  

<br/>

--

#### A random effects model is appropriate when the treatment levels in the experiment can be considered a sample from a larger population of interest  

- In the previous example, we are interested in all instructors, not just the ones included in the experiment  


---
# models

### Fixed-effects model  

$$\Large y_{ij} = \mu + \alpha_i + \epsilon_{ij}$$
$$\Large  \epsilon_{ij} \simnormal(0, \sigma^2)$$
--

### Random-effects model

$$\Large y_{ij} = \mu + \alpha_i + \epsilon_{ij}$$
$$\Large \alpha_{i} \simnormal(0, \sigma_A^2)$$

$$\Large \epsilon_{ij} \simnormal(0, \sigma^2)$$

--

In a random effects model, the effects are assumed to be random variables following some probability distribution

---
# hypotheses

#### In a random-effects model, our interest is in assessing how much variation there is among all the effects in the population, not just the ones in our sample  


#### As a result, our hypotheses must be written differently:

--

#### Fixed-effects model

$$\large H_0 : \alpha_1 = \alpha_2 = ... = \alpha_a = 0$$

$$\large H_a : At\;least\;one\;inequality$$

--

#### Random-effects model

$$\large H_0 : \sigma^2_A = 0$$


$$\large H_a : \sigma^2_A > 0$$


---
# implications?

#### Oddly enough, in the context of a one-way ANOVA, the analysis procedure is exactly the same for the fixed-effects and random-effects models  

--

#### In other words, we can construct the ANOVA table and test the null hypothesis using the same procedure for the random-effects model as we used for the fixed-effects model, even though the interpretation of the results is slightly different  

--

#### A comprehensive list of the distinctions between the two approaches is given in section 11.1 of Dowdy et al.

--

#### The distinction between random and fixed effects becomes much more important in the more complicated models that we will cover soon  

---
class: inverse, center, middle

# random effects in blocked anova

---
# blocked anova

<br/>

### Should we treat block effects as fixed or random?  

<br/>

--

### As before, we need to answer the question: 

> Can we view the blocks as samples of a larger population, or do the blocks represent all possible levels?


---
# random effects model for blocked design


$$\huge y_{ij} = \mu + \alpha_i + \beta_j + \epsilon_{ij}$$
$$\Large \beta_{j} \sim normal(0, \sigma^2_B)$$

$$\Large \epsilon_{ij} \sim normal(0, \sigma^2)$$

<br/>

--

#### This is often called a mixed effects model because it includes fixed and random effects  

--

#### If we treated the $\alpha$'s as random too, we might call the model a *variance components model*  



---
# hypotheses

.pull-left[
#### Main hypothesis

$$H_0 : \alpha_1 = \alpha_2 = ... = \alpha_a = 0$$

$$H_a : At\;least\;one\;inequality$$

OR

$$H_0 : \sigma^2_A = 0$$

$$H_a : \sigma^2_A > 0$$
]
--

.pull-right[

#### Secondary hypothesis

$$H_0 : \beta_1 = \beta_2 = ... = \beta_b = 0$$

$$H_a : At\;least\;one\;inequality$$
OR

$$H_0 : \sigma^2_B = 0$$

$$H_a : \sigma^2_B > 0$$
]

---
# implications

<br/>
<br/>

#### Once again, treating block effects as random doesn't affect our calculations for the ANOVA table<sup>1</sup>  - it only affects the interpretation


.footnote[<sup>1</sup> Technically, this is only true for balanced designs]

---
# random-effects model in `r`

```{r echo = FALSE}
gypsyData <- data.frame(location = rep(c("A", "B", "C", "D"), 3),
                        caterpillars = c(16, 3, 10, 18, 25, 10, 15, 32, 14, 2, 16, 12),
                        pesticide = rep(c("Bt", "Control", "Dimilin"), each = 4))
```

```{r echo = TRUE, eval = FALSE}
aov1 <- aov(caterpillars ~ pesticide + location, gypsyData)
summary(aov1)
```

```{r echo = FALSE}
aov1 <- aov(caterpillars ~ pesticide + location, gypsyData)
options(knitr.kable.NA = '')
kable(broom::tidy(aov1)) %>%
  kable_styling(font_size = 10, full_width = FALSE, bootstrap_options = "condensed")
```


--

```{r echo = TRUE, eval = FALSE}
aov2 <- aov(caterpillars ~ pesticide + Error(location), gypsyData)
summary(aov2)
```

```{r echo = FALSE}
aov2 <- aov(caterpillars ~ pesticide + Error(location), gypsyData)
options(knitr.kable.NA = '')
kable(broom::tidy(aov2)) %>%
  kable_styling(font_size = 10, full_width = FALSE, bootstrap_options = "condensed")
```



---
# summary

- We usually model the main treatment effects as fixed, but sometimes we are interested in effects not in the sample

--

- Random effects models allow for inference about entire population of effects  

--

- Random effects models make additional distributional assumptions  

--

- Usually, you need many treatment levels (5-10) to get a decent estimate of the variance parameter associated with the random effects  

--

- In spite of the big conceptual differences, the procedures aren't much different for one-way and blocked ANOVAs  



