---
title: "LECTURE 9: non-parametric statistics"
subtitle: "FANR 6750 (Experimental design)"
author: "<br/><br/><br/>Fall 2021"
output:
  xaringan::moon_reader:
    css: ["default", "FANR6750.css", "FANR6750-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, echo = FALSE, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning=FALSE, message=FALSE, 
                      fig.retina = 2, fig.width = 9, fig.height = 7)
library(WILD3810)
library(emo)
library(kableExtra)
library(dplyr)
library(FANR6750)
# library(gganimate)
```

# outline 

<br/>

1) Motivation

<br/>

--

2) Kruskal-Wallis test

<br/>

--

3) Wilcoxan test


---
# motivation

#### Parametric models assume that random variables follow standard probability distributions, such as:

- normal (Gaussian)  

- gamma  

- binomial  

- Poisson  

- multinomial  

--

#### The model underlying ANOVA is parametric because it assumes: 

$$\large \epsilon_{ij} \sim normal(0, \sigma^2)$$

---
# what if assumptions don't hold?

### At least three options:

1) Use a more complicated model that relaxes some assumptions, such as:

  + Generalized linear models that don't assume normality  

  + Linear models with multiple variance parameters  

  + Time-series models or spatial models allowing for correlated residuals  



**2) Non-parametrics**  


3) Transform the data

<br/>

--

#### Non-parametric statistics are distribution free (but not assumption free!)

---
# non-parametric statistics

#### Key idea

> Replace the original observations with their ranks, which will result in tests that are less sensitive to extreme values  

--

```{r}
rank_df <- data.frame(Group = rep(c("A", "B"), each = 4),
                      Original = c(9, 3, 11, 4, 10, 992, 8, 2),
                      Rank = c(5, 2, 7, 3, 6, 8, 4, 1))

rank_df %>%
  kable(format = 'html', col.names = c("Group", "Original data", "Rank-transformed data"), 
        align = 'c') %>%
  kable_styling(font_size = 11, bootstrap_options = c("condensed"), full_width = FALSE)
```

---
class: inverse, center, middle

# kruskal-wallis test

---
# kruskal-wallis test

#### This is the non-parametric counterpart to one-way ANOVA

--

### Procedure

1) Rank the data  

--
2) Compute the test statistic:  

$$\large H = \frac{n\big[\sum_i(\bar{r}_i - (N+1)/2)^2\big]}{N(N+1)/12}$$
where $\bar{r}_i$ is the mean rank of group $i$, and $N$ is the total number of observations across all the groups.

--

3) Compare $H$ to a chi-squared distribution and reject null hypothesis if $H \geq \chi^2_{\alpha,a-1}$

---
# $\chi^2$ distribution

```{r}
x <- seq(from = 0, to = 10, by = 0.1)

chi2 <- data.frame(x = x,
                   y = dchisq(x, df = 3))

ggplot(chi2, aes(x = x, y = y)) +
  geom_path() +
  scale_x_continuous(expression(chi^2)) +
  scale_y_continuous("") +
  labs(title = "Chi−squared distribution with df=3")
```

---
# but what is the null hypothesis?

#### Null hypothesis

> The mean rank of all groups is equal to the mean rank of the entire data set

$$\large H_0 : E(\bar{r}_i) = \frac{N + 1}{2}$$
for all $i$  

--
#### Alternative hypothesis

> The mean rank of at least one group is not equal to the mean rank of the entire data set

$$\large H_a : E(\bar{r}_i) \neq \frac{N + 1}{2}$$

for at least one $i$

---
# back to the chainsaw data

```{r}
cs2 <- data.frame(blank = c(NA, NA, NA, NA, NA),
                  degree1 = c(42, 17, 24, 39, 43),
                  rank1 = c(12, 1, 3, 9, 13),
                  degree2 = c(28, 50, 44, 32, 61),
                  rank2 = c(4, 17, 14, 7, 20),
                  degree3 = c(57, 45, 48, 41, 54),
                  rank3 = c(19, 15, 16, 11, 18),
                  degree4 = c(29, 40, 22, 34, 30),
                  rank4 = c(5, 10, 2, 8, 6))
options(knitr.kable.NA = '')
cs2 %>%
  kable(format = 'html', col.names = c("", "Degrees", "Rank",
                                           "Degrees", "Rank", 
                                           "Degrees", "Rank", 
                                           "Degrees", "Rank")) %>%
  add_header_above(c("Model" = 1, "A" = 2, "B" = 2, "C" = 2, "D" = 2)) %>%
  kable_styling(full_width = TRUE, bootstrap_options = "condensed", font_size = 12)
```

--

```{r}
r.i <- data.frame(model = '\\(\\bar{r}_i\\)',
                  degree1 = NA,
                  rank1 = "7.6",
                  degree2 = NA,
                  rank2 = "12.4",
                  degree3 = NA,
                  rank3 = "15.8",
                  degree4 = NA,
                  rank4 = "6.2")

r.i %>%
  kable(format = 'html', col.names = NULL, escape = TRUE, align = "r") %>%
  kable_styling(full_width = TRUE, bootstrap_options = "condensed", font_size = 12)
```

--

$$\large H = \frac{293}{35} = 8.37$$

--

Reject the null because $8.37$ is greater than $\chi^2_{0.05;3} = 7.815$

---
class: inverse, center, middle

# wilcoxon rank sum test

---
# wilcoxon rank sum test

#### Suppose we just wanted to compare models A and D  

--

#### A non-parametric alternative to the two-sample $t$ test is the Wilcoxon rank sum test

- This is also known as the Mann-Whitney U test  

- For paired data, there is the the Wilcoxon signed-rank test  

--

#### The procedure is similar as before except the test statistic is different and we compare it to a standard normal distribution  


---
# wilcoxon rank sum test

### Test statistic

$$\Large z = \frac{\bar{r}_1 - (N + 1)/2}{\sqrt{(N - n_1)(N + 1)/(12n_1)}}$$

--

### Decision rule

- Reject if $|z| > z_{\alpha/2}$ (for a two-tailed test)  

- $z_{\alpha/2}$ is for the standard normal distribution, i.e. $normal(0, 1)$

---
# summary

<br/>

--

- Nonparametric tests are simple and easy to implement  

--

- However, they aren't as efficient as their parametric counterparts  

--

- They are also much harder to devise for complicated models such as those that we will cover soon  

--

- Parametric assumptions usually are not problematic
