---
title: "LECTURE 9: assumptions and transformations"
subtitle: "FANR 6750 (Experimental design)"
author: "<br/><br/><br/>Fall 2022"
output:
  xaringan::moon_reader:
    css: ["default", "FANR6750.css", "FANR6750-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, echo = FALSE, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning=FALSE, message=FALSE, 
                      fig.retina = 2, fig.width = 9, fig.height = 7)
source(here::here("R/zzz.R"))
library(emo)
library(kableExtra)
library(dplyr)
library(FANR6750)
# library(gganimate)
```

# outline 

<br/>

1) Motivation

<br/>

--

2) Assumptions

<br/>

--

3) Transformations


---
# assumptions

The key assumptions of ANOVA are that the residuals are **independent** and come from a **normal distribution** with mean 0 and **variance** $\sigma^2$  

--

#### This implies the following:

1) Within each treatment group, the residuals are normally distributed  

--

2) Treatment group variances are equal  

--

3) There is no spatial, temporal, or other forms of correlation among the residuals - they are independent

---
# normality assumption

<br/>

### Assessment

--
- Inspect histograms and boxplots of the data for each treatment group  

--

- Inspect histograms and boxplots of the residuals  

--

- Conduct a Shapiro-Wilk test of normality on the residuals

    + Goodness-of-fit test of normality  
    
    + Assumption of normality is rejected if p < 0.05 

---
# normality assumption

```{r}
data("warblerWeight")
warblerWeight2 <- dplyr::filter(warblerWeight, population %in% c("A", "B", "C", "D"))

ggplot(warblerWeight2) +
     geom_histogram(aes(weight), binwidth = 1, color = "black", fill = "grey70") +
     facet_wrap(.~population, nrow = 2, ncol = 2) +
  scale_x_continuous("Warbler weight") +
  scale_y_continuous("Frequency")
```

---
# normality assumption

#### ANOVA is usually considered robust to violations of this assumption as long as they aren't severe and the sample size isn't too small.  

#### However, some problems can arise  

<br/>

--

**Consequences of non-normality**  

- Implausible confidence intervals  

- Altered power if the data are heavily skewed (platykurtic or leptokurtic)

---
# normality assumption

<br/>

#### Cases where normality assumption might not hold  

- Count data, especially when there are lots of zeros  

- Proportion or percentage data  

- Arbitrary scales, such as a 10-point taste test  

- Weights of very small things  

---
# equal variance assumption

<br/>

### Assessment

--
- Inspect histograms and boxplots of the data from each treatment group

--

- Plot group variances  

--

- Conduct a Bartlett test of equality of variances  

    + Assumption of homogeneity is rejected if p < 0.05

---
# equal variance assumption

```{r}
set.seed(340)
x <- c(rpois(50, 0.5), rpois(50, 2), rpois(50, 8))
ff <- gl(3, 50, labels=c("Pine", "Oak-Hickory", "Prairie"))

deer_df <- data.frame(Abundance = x, 
                      Habitat = ff)

ggplot(deer_df) +
     geom_histogram(aes(Abundance), binwidth = 1, color = "black", fill = "grey70") +
     facet_wrap(.~Habitat, nrow = 1) +
  scale_x_continuous("Deer abundance") +
  scale_y_continuous("Frequency")
```


---
# equal variance assumption

```{r}
library(dplyr)
var.i <- deer_df %>%
  group_by(Habitat) %>%
  summarise(mu = mean(Abundance),
            alpha.i = mu - mean(deer_df$Abundance),
            var.i = var(Abundance))

ggplot(var.i, aes(x = Habitat, y = var.i)) +
  geom_col(color = "black", fill = "grey70") +
  scale_y_continuous("Variance")
```




---
# equal variance assumption

#### ANOVA is again considered robust to violations of this assumption, especially if sample sizes are similar among groups.

--
#### Consequences of heteroscedasticity

--

- Estimates of variance ( $\sigma^2$ ) will be wrong  

- Confidence intervals and F-tests will be affected  

--

#### Cases where heteroscedasticity might be expected

- Count data, especially when there are lots of zeros  

- Proportion or percentage data  

- Arbitrary scales, such as a 10-point taste test

---
# independence assumption

<br/>

--

#### Residuals are independent if the value of one residual tells us nothing about the value of another residual  

<br/>

--

#### This assumption can be met by randomly selecting experimental units and randomly assigning them to treatment groups  

<br/>

--

#### It is unlikely to hold if multiple observations are made on the same experimental unit, or if the experimental units are highly clustered  

---
# independence assumption

### Assessment

--

- In general, this problem is hard to diagnose unless you know the details of the design  

--

- Plot group variances against group means. There shouldn't be any pattern  

--

- When temporal or spatial autocorrelation is a concern, auto-covariance plots can be used for diagnosis  

--

### Consequences of non-independence

- Estimates of variance will generally be too small

- Power will be in inflated

---
# what if assumptions don't hold?

### At least three options:

--
1) Use a more complicated model that relaxes some assumptions, such as:

  + Generalized linear models that don't assume normality  

  + Linear models with multiple variance parameters  

  + Time-series models or spatial models allowing for correlated residuals  


--

2) Nonparametrics  

<br/>
--

3) Transform the data

---
class: inverse, center, middle

# transformations

---
# transformations

### Key idea

> Transform the raw data ( $u_{ij}$ ) so that the transformed data ( $y_{ij}$ ) meet the assumptions of ANOVA  

--

#### What kinds of transformations are valid?

- The transformation must maintain the rank order of the original data

- Common examples include:

    + Logarithmic transformations  

    + Square-root transformations  

    + Arcsine transformations  
    
    + Reciprocal transformations

---
# logarithmic transformation

<br/>

$$\huge y = log(u + C)$$

<br/>

--

- The constant C is often 1, or 0 if there are no zeros in the data ( $u$ )

--

- Useful when group variances are proportional to the means (count data)

--
- Could use $log_{10}$ or any base, but the natural logarithm is preferred

---
# logarithmic transformation

```{r}
# Weights
u1 <- data.frame(u = c(4.4, 5, 5.6, 4.2, 4.3, 5.3, 4.2, 3.6, 4.8, 4.9,
        5.7, 4.6, 5.8, 6.3, 4.5, 4.1, 3.6, 4, 4.5, 4.6,
        6.6, 4.3, 4.8, 3.7, 5.5, 2.9, 4.3, 4.2, 3.6, 4.7))
#
#
# Tree diameters
u2 <- data.frame(u = c(137.6, 102.6, 327.6, 550.4, 214, 86, 185.2, 24.5,
        134.6, 169.1, 263.5, 226.4, 225.3, 82, 391.8, 567.1,
        74.8, 52.8, 72.2, 474.2, 138, 42.4, 172.5, 209.6,
        102.6, 563.4, 241.1, 260.6, 203.1, 137.9))

ggplot(u2, aes(x = u)) +
  geom_histogram(binwidth = 100, 
                 fill = "grey70", color = "black") +
  scale_y_continuous("Frequency")

```

---
# logarithmic transformation

```{r}

ggplot(u2, aes(x = log(u))) +
  geom_histogram(binwidth = 0.4, fill = "wheat", color = "black") +
  scale_y_continuous("Frequency") +
  scale_x_continuous("y = log(u)")

```

---
# square root transformation

<br/>
$$\huge y = \sqrt{u + C}$$

<br/>

--

- C is often 0.5 or some other small number

--

- Useful when group variances are proportional to the means (count data)

---
# square root transformation

```{r}

ggplot(u2, aes(x = u)) +
  geom_histogram(binwidth = 100, 
                 fill = "grey70", color = "black") +
  scale_y_continuous("Frequency")

```

---
# square root transformation

```{r}
ggplot(u2, aes(x = sqrt(u))) +
  geom_histogram(binwidth = 4, fill = "wheat", color = "black") +
  scale_y_continuous("Frequency") +
  scale_x_continuous("y = sqrt(u)")
```



---
# summary

--
- The assumptions of ANOVA are made explicit by the linear model  

--

- Colloquially, these assumptions are:  

    + Within each treatment group, the data are normally distributed  
    
    + Treatment group variances are equal  
    
    + There is no spatial, temporal, or other forms of correlation among the residuals - they are independent  
    
--

- ANOVA is considered robust to violations of the first two assumptions  

--

- Transformations, in some cases, can be used to meet all three assumptions

--
- If transformations don't work, it is always possible to relax assumptions by adopting a more complicated model
